Labor Market Income and Substitution Effects Explain Modern Natural Selection

### 2020-12-16 why-negative-selection.Rmd
Abdel's comments, restyling

---
title: "Labor Market Income and Substitution Effects Explain Modern Natural Selection"
author: "David Hugh-Jones (D.Hugh-Jones@uea.ac.uk), Abdel Abdellaoui (a.abdellaoui@gmail.com)"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection on 33 polygenic scores over two generations, using data from UK Biobank. Consistently over time, polygenic scores which predict lower (higher) earnings and education are selected for (against). Selection effects are concentrated among low-income and less educated people, younger parents, people with more lifetime sexual partners, and people not living with a partner. Effect sizes are substantially larger after correcting for ascertainment bias. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns can be explained by economic theories of fertility, in which higher earnings may either increase or decrease fertility via income and substitution effects in the labour market."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Times
---

@@

As yet we know little about the mechanisms behind these effects. This study uses
data from UK Biobank to learn more. We test for natural selection on 33
different polygenic scores by estimating their correlation with fertility. We
extend the analysis over two generations, using data on respondents' number of
siblings as well as their number of children. This is interesting because
consistent natural selection over multiple generations could lead to substantive
effects in the long run. Most importantly, we examine reproductive rates in
different subgroups of the population, in order to uncover patterns that can
help illuminate the mechanisms behind modern natural selection.

We find selection effects on many polygenic scores. Effects are consistent
across generations. The strength of natural selection on a polygenic score is
associated with that score's correlation with education and earnings: scores
that predict lower education and earnings are being selected for. Also, across
the board, polygenic scores have stronger relationships with fertility among
specific subgroups. Selection effects are stronger among people with lower
income and less education, younger parents, people not living with a partner,
and people with many lifetime sexual partners. Outside these groups, effects are
weaker and often statistically insignificant. In fact, in some subgroups, the
direction of selection is actually reversed: polygenic scores predicting
higher education and earnings are associated with *higher* fertility.

These patterns can be explained by economic theories of fertility
[@becker1960economic]. In these, higher potential earnings have two opposite
effects on fertility: a fertility-increasing *income effect* (higher income
makes children more affordable), and a fertility-lowering *substitution effect*
(time spent on childrearing has a higher cost in foregone earnings). Our
results suggest that the substitution effect dominates for single parents,
while the income effect is stronger for couples.

@@

Figure \@ref(fig:plot-means-over-time) plots mean polygenic scores in the
sample by 5-year birth intervals. Several scores show consistent increases or
declines over this 30-year period, of the order of 5% of a standard
deviation.These changes could reflect natural selection within the UK
population, but also ascertainment bias within the sample. Respondents are
higher income and better educated than the UK population, and they may
also differ on other unobserved characteristics [@10.1093/aje/kwx246].
Since richer and educated people also live longer, this bias might also
increase with age.

```{r plot-means-over-time, fig.height = 6, fig.cap = "Mean polygenic scores by birth year in UK Biobank. Points are means for 5-year intervals. Lines are 95% confidence intervals. Green triangles show a significant linear increase over time (p < 0.05/33). Red squares show a significant decrease."}

@@

To test directly for natural selection, we regress respondents'polygenic scores
on their number of children ($y_i$):

@@

The "selection effect", $\beta$, measures the strength of natural selection
within the sample. In fact, since polygenic scores are normalized, $\beta$ is
the expected polygenic score among children of the sample
[@beauchamp2016genetic].[^normalization] To learn more about the underlying
mechanisms, we split the sample, starting with basic demographic variables
including education, income and sex. These are all potential sources of
ascertainment bias: as well as the ascertainment for income and education,
mentioned above, the sample sex ratio skews `r female_ratio * 100`% female.

@@

These results could be driven by age, if older respondents are poorer
and less educated, and also more subject to selection on polygenic
scores. However, if we rerun the regressions, interacting the polygenic
score with income category and also with a quadratic in age, the
interaction with income remains significant at 0.05/`r n_regs_ic` for
`r n_sig_ic` out of `r n_regs_ic` regressions. Similarly if we interact
the PGS with age of leaving full time education and a quadratic in age,
the interaction with age leaving FTE remains significant at
0.05/`r n_regs_educ` for `r n_sig_educ` out of `r n_regs_educ` regressions.

Selection effects are also different between men and women (Appendix
Figure \@ref(fig:plot-sexes)). Differences are particularly large for
educational attainment, height and MDD. Several PGS for mental illness
and personality traits are more selected for (or less against) among
women, including major depressive disorder (MDD), schizophrenia and
neuroticism, while extraversion is more selected for among men. PGS for waist
circumference and waist-hip ratio are less selected for among women, and PGS for
educational attainment are more selected against, though the difference
is only significant for EA3. One possible reason for these sex differences is
that polygenic scores may affect fertility via success in marriage markets, and
men and women may value different characteristics in these.

@@

We next focus on variables related to household type and reproductive
strategy. We split males and females by lifetime number of sexual
partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
For both sexes, selection effects are larger and more significant among
those with more than 3 partners. Next we split respondents by whether
they were living with a spouse or partner at the time of interview.
Effects are larger among those not living with a spouse or partner
(Figure \@ref(fig:plot-n-partners)b). Lastly, we split female
respondents by age at first live birth (AFLB).[^1] There is evidence for
genetic effects on AFLB [@Barban_2016], and there is a close link
between this variable and number of children born. Figure
\@ref(fig:plot-n-partners)c shows effect sizes estimated separately for
each tercile of AFLB. Several effects are strikingly different across
terciles. ADHD and MDD are selected for amongst the youngest third of
mothers, but selected against among the oldest two-thirds. Educational
attainment is selected for among the oldest two-thirds of mothers, but
is not significantly selected among the youngest third. Similarly,
several PGS of body measurements are selected against only among older
mothers. The correlation between effect sizes for the youngest and
oldest terciles is `r cor_age_flb`. As before, recall that PGS may have
two effects: changing the number of children within each category, and changing
the number of children by shifting respondents between categories.

@@

@@

@@

To investigate this further, we regress number of children on PGS
*controlling* for AFLB (appendix Figure \@ref(fig:plot-age-flb)). In
`r opp_signed` out of `r length(score_names)` cases, effects
change sign when controls are added. The correlation between effect
sizes controlling for AFLB, and raw effect sizes, is `r cor_raw_flb`.
Thus, selection effects seem to come through two opposing channels: an
effect on AFLB, and an opposite-signed effect on number of children 
controlling for AFLB.

@@

n_regs <- length(score_names)
res_wt_combined %>% 
  mutate(
    Weights  = fct_relevel(Weights, "Geographical", "Age/Qualification",
                "Age/Qual/AFLB", "None"),
    term     = pretty_names(term),
    term     = fct_reorder(term, estimate, order_abs(1))
  ) %>% 
  rename(p = p.value) %>% 
  ggplot(aes(
    x = estimate, 
    y = term, 
    shape = Weights, 
    fill = Weights, 
    colour = p < 0.05/{{n_regs}}
  )) +
  geom_point(size = 1.5, stroke = 0.5) +
  my_vline +
  pval_scale() +
  scale_fill_manual(values = c(
    "None"              = "grey15", 
    "Geographical"      = "dodgerblue4",
    "Age/Qualification" = "steelblue2",
    "Age/Qual/AFLB"      = "darkseagreen4"
  )) +
  scale_shape_manual(values = c(
    "None" = "square filled", 
    "Geographical" = "circle filled",
    "Age/Qualification" = "circle filled", 
    "Age/Qual/AFLB" = "circle filled"
  )) + 
  labs(x = "Effect size", y = "")

@@

The UK Biobank data contains information on respondents' number of
siblings. Since respondents' polygenic scores are equal in expectation
to the mean scores of their parents, we can use this to look at selection 
effects in the parents' generation. We estimate equation (\ref{eq:regression})
using *number of siblings* (including the respondent) as the dependent variable.
The parents' generation has a further source of ascertainment bias: sampling
parents of respondents overweights parents who have many children. For instance,
parents of three children will have, on average, three times more children
represented in UK Biobank than parents of one child. Parents of no children will
by definition not be represented. To compensate, we reweight our preferred
weightings (Age/Qualification) by the inverse of *number of siblings*.

Selection effects are highly correlated across the two generations, and most
share the same sign (appendix Figure \@ref(fig:plot-res-sibs)). Effect size
estimates are larger for the parents' generation. We treat this result cautiously,
because when we split respondents up by year of birth, we find few differences
in effect sizes between early- and late-born respondents, for either generation.
In other words, since estimated effect sizes change when we change the dependent
variable, but do not change over time within either dependent variable, this
may be due to remaining ascertainment bias within the sample.

Although the direction of selection effects does not change between the
generations, there are other differences. Compared to residualized polygenic
scores, selection effects on unresidualized scores are about ten percent higher
on average for *n siblings*, whereas effects for *number of children* barely
change (appendix Figure \@ref(fig:plot-regs-controlled-pcs)). This could be
because earlier fertility is driven more by geographically clustered deprivation
[e.g. via an insurance motive, @rendall1998old], which may correlate with the
broad-scale genetic variation captured by principal components.

We also check whether selection effects differ by socio-economic status in the
parents' generation. We have no information about parents' income, so we use the
1971 Townsend deprivation score of respondents' birthplace as a proxy
[@townsend1987deprivation]. Results (appendix Figure
\@ref(fig:plot-siblings-townsend)) show the same pattern as for respondents:
effect sizes are larger and more often significant in the most deprived areas.

```{r calc-siblings-children}
sib_chn_cor <- cor(famhist$n_sibs, famhist$n_children, use = "complete")
```

Lastly, the siblings data lets us check for a "quantity-quality tradeoff" between
number of children and number of grandchildren (for the parents' generation). We
do not find any: in fact, the correlation between *number of siblings* and 
*number of children* is positive ($\rho = `r sib_chn_cor`$).

@@

These results show that selection effects are weaker, absent, or even reversed
among some subgroups of the population. This can potentially be explained by the
economic theory of fertility [@becker1960economic; @willis1973new;
@becker1976child]. According to this, increases in a person's wage affect their
fertility via two opposing channels. There is an *income effect* by which
children become more affordable, like any other good. There is also a
*substitution effect*: since childrearing has a cost in time, the opportunity
cost of childrearing increases if one's market wage is higher. The income effect
would lead higher earners to have more children. The substitution effect would
lead higher earners to have fewer children. If so, then genetic variants which
affect earnings potential in the labour market may cause opposing effects on
fertility. The income effect will cause natural selection in favour of
earnings-increasing variants. The substitution effect will do the reverse.

@@

The theory can also explain why natural selection is weaker among higher-income
parents. @becker1976child show that if parents care about child quality as well
as quantity, under certain assumptions the income effect will be stronger at
higher income levels; this can lead to a U-shaped relationship between income
and fertility. Empirically, @cohen2013financial find that income decreases
fertility at low income levels but increases it at higher income levels. If, in
our data, at high income levels income and substitution effects are roughly
balanced , then income-linked polygenic scores will be neither selected for nor
against.

@@

Second, we re-estimate equation \@ref(eq:regression) for each
PGS, controlling for earnings in first job and education levels. Effect sizes
are generally reduced (appendix Figure \@ref(fig:plot-res-earn-educ)), and only
`r n_05_33` out of 33 PGS are significant at $p = 0.05/33$. This suggests that
earnings and education indeed mediate the effect of PGS on fertility.

@@

An alternative theory is that traits selected for are linked to externalizing
behaviour, risk-seeking and low time discount rates, via the channel of early
sexual behaviour [@Mills2020.05.06.081273]. The data here provide some support
for this: scores which might plausibly be linked to externalizing behaviour,
like ADHD and age at smoking initiation, are selected for. However, this theory
is less good at explaining variation in selection across the full range of
scores, including physical measures, e.g. waist-hip ratio and BMI. Externalizing
behaviour also does not explain why selection effects should work in the
opposite direction among older parents, whereas the economic theory can explain
this via income effects, as described above. We test the alternative theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (Biobank field 2040). While risk attitude is always a
highly significant predictor of number of children, it has little impact on the
effect size or significance of the PGS. The median ratio of PGS effect sizes
between regressions with and without controls is `r median(ratio)`; all PGS
which are significant at $p \le 0.05/33$ in uncontrolled regressions remain so
when controlling for risk attitude. Thus, although risk attitude does predict
number of children, it appears not to mediate selection effects. Overall, we
believe that the economic theory is the most likely explanation.

@@

Previous work has documented natural selection in modern populations on variants
underlying many polygenic scores [@beauchamp2016genetic; @kong2017selection;
@sanjak2018evidence]. We show that correlations between polygenic scores and
fertility are highly concentrated among specific subgroups of the population,
including poorer people, less educated people, younger parents and those with
more lifetime sexual partners. Indeed, among older parents selection effects are
actually reversed. Furthermore, the size of selection effects on a polygenic
score correlates with that score's association with labour market earnings. The
economic theory of fertility provides a parsimonious explanation for this.

PGS which correlate with high (low) earnings and more (less) education are being
selected against (selected for). In addition, many of the phenotypes under
positive selection are linked to poor health, or are what many people would see
as undesirable to have. For example, most people would probably prefer to have
high educational attainment, a low risk of major depressive disorder, and a low
risk of coronary artery disease, but natural selection is pushing against
genetics linked to these traits. Potentially, this could increase the health
burden on modern populations, but that depends on effect sizes. Our results
suggest that naïve estimates can be affected by sample ascertainment bias. This
problem may be less serious in surveys which aim to be representative (as the UK
Biobank does not). However, there is still scope for bias, since not all
respondents consent to the collection of genetic data. For instance, completion
rates for genotype data in the US Health and Retirement Study were around 80-85%
[@hrs2020]. Researchers should be aware of the risks of ascertainment when
studying modern natural selection.

We also do not know how estimated effect sizes of natural selection will change
as more accurate measures of genetic variation are produced. And we are unsure
whether genetic variants underlying other phenotypes will show a similar pattern
of natural selection to those studied here. In short, it is probably too early
to tell whether modern natural selection has a substantively important effect on
population genetics. Nevertheless, selection effects on our measured
polygenic scores are still relatively small, even after reweighting to 
account for ascertainment bias.

Because selection effects are concentrated in poorer groups, they also increase
inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
attainment (EA3), for children in households of different income groups. The
blue bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against EA3, which lowers the
mean, is stronger at the bottom of the income distribution, and this increases
the differences between groups.

@@

Any model of fertility is implicitly a model of natural selection. But so far,
the economic and human genetics literatures have developed in parallel.
Integrating the two could deepen our understanding of modern natural selection.
Economics possesses a range of theoretical models on the effects of skills,
education and income [see @hotz1997economics, @lundberg2007american]. One
perennial problem is how to test these theories in a world where education,
labour and marriage markets all interact. Genetic data, such as polygenic
scores, could help to pin down causality. Conversely, theory and empirical
results from economics can shine a light on the mechanisms behind natural
selection.

@@

We use participant data from UK Biobank [@sudlow2015uk], which has received
ethical approval from the National Health Service North
West Centre for Research Ethics Committee (reference: 11/NW/0382).
We limit the sample to white British participants, as defined by
genetic estimated ancestry and self-identified ethnic group, giving a sample
size of `r comma_num(nrow(famhist))`. For regressions on number of children we use participants over 45, since most fertility is completed by this age. This
gives a sample size of `r comma_num(nrow(famhist %>% filter(kids_ss)))`.

We computed 33 polygenic scores. The effect size estimates were obtained from
GWASs that were chosen to not have included the UK Biobank dataset, to avoid
overestimation of the genetic predisposition of a trait. The polygenic scores
were computed using the summary-data based best linear unbiased prediction
(SBLUP) approach. As a reference sample for the linkage disequilibrium, we used
a random sample of 10,000 unrelated individuals from UK Biobank, imputed using
the Haplotype Reference Consortium panel. Scores were normalized to mean 0,
variance 1, and residualized on the first 100 principal components of array
data, which were calculated as in @abdellaoui2019genetic.

@@

\FloatBarrier
\newpage
# Appendix

@@

\FloatBarrier
## Controlling for AFLB

```{r plot-age-flb, fig.cap = "Selection effects controlling for age at first live birth (women only). Effect sizes for women only without controls are shown for comparison."}

>>>>

---
title: "Labor Market Income and Substitution Effects Explain Modern Natural Selection"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Amsterdam UMC. Email: a.abdellaoui@amsterdamumc.nl}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection on 33 polygenic scores over two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores which predict lower (higher) earnings and education are selected for (against). Selection effects are concentrated among low-income and less educated people, younger parents, people with more lifetime sexual partners, and people not living with a partner. Effect sizes are substantially larger after correcting for ascertainment bias. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns can be explained by economic theories of fertility, in which higher earnings may either increase or decrease fertility via income and substitution effects in the labour market. We find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects and that are strongly related to the socio-economic system."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\bfseries\sffamily\LARGE}
  - \titleformat*{\subsection}{\bfseries\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Times
---

@@

As yet we know little about the social mechanisms behind these effects. This study uses
data from UK Biobank [@bycroft2018uk] to learn more. We test for natural selection on 33
different polygenic scores by estimating their correlation with fertility. We
extend the analysis over two generations, using data on respondents' number of
siblings as well as their number of children. This is interesting because
consistent natural selection over multiple generations could lead to substantive
effects in the long run. Most importantly, we examine reproductive rates in
different subgroups of the population, in order to uncover patterns that can
help illuminate the mechanisms behind modern natural selection.

We find selection effects on many polygenic scores. Effects are largely consistent
across generations. The strength of natural selection on a polygenic score is
associated with that score's correlation with education and earnings: scores
that predict lower education and earnings are being selected for. Also, across
the board, polygenic scores have stronger relationships with fertility among
specific subgroups. Selection effects are stronger in groups with lower
income and less education, among younger parents, people not living with a partner,
and people with many lifetime sexual partners. Outside these groups, effects are
weaker and often statistically insignificant. In some subgroups, the
direction of selection is even reversed: polygenic scores predicting
higher education and earnings are associated with *higher* fertility.

These patterns are in line with economic theories of fertility
[@becker1960economic]. In these, higher potential earnings have two opposite
effects on fertility: a fertility-increasing *income effect* (higher income
makes children more affordable), and a fertility-lowering *substitution effect*
(time spent on childrearing has a higher cost in foregone earnings). Our
results suggest that the substitution effect dominates for single parents,
while the income effect is stronger for couples.

@@

Figure \@ref(fig:plot-means-over-time) plots mean polygenic scores in the
sample by 5-year birth intervals. Several scores show consistent increases or
declines over this 30-year period, of the order of 5% of a standard
deviation. These changes could reflect natural selection within the UK
population, but also emigration, or ascertainment bias within the sample. 
Respondents have higher income and are better educated than the UK population, 
and they may also differ on other unobserved characteristics [@10.1093/aje/kwx246].
Since richer and educated people also live longer, this bias could also
increase with age.

```{r plot-means-over-time, fig.height = 6, fig.cap = "Mean polygenic scores (PGS) by birth year in UK Biobank. Points are means for 5-year intervals. Lines are 95% confidence intervals. Green triangles show a significant linear increase over time (p < 0.05/33). Red squares show a significant decrease."}

@@

To test directly for natural selection, we regress respondents' polygenic scores
(PGS) on their number of children ($y_i$):

@@

The "selection effect", $\beta$, reflects the strength of natural selection
within the sample. In fact, since polygenic scores are normalized, $\beta$ is
the expected polygenic score among children of the sample
[@beauchamp2016genetic].[^normalization] To learn more about the underlying
mechanisms, we split the sample, starting with basic demographic variables
including education, income and sex. These are all potential sources of
ascertainment bias: as well as the ascertainment for income and education,
mentioned above, the sample sex ratio skews `r female_ratio * 100`% female.

@@

These results could be driven by age, if older respondents have lower income
and are less educated, and also more subject to selection on polygenic
scores. However, if we rerun the regressions, interacting the polygenic
score with income category and also with a quadratic in age, the
interaction with income remains significant at 0.05/`r n_regs_ic` for
`r n_sig_ic` out of `r n_regs_ic` regressions. Similarly if we interact
the PGS with age of leaving full time education and a quadratic in age,
the interaction with age leaving FTE remains significant at
0.05/`r n_regs_educ` for `r n_sig_educ` out of `r n_regs_educ` regressions.

Selection effects are also different between men and women (Appendix
Figure \@ref(fig:plot-sexes)). Differences are particularly large for
educational attainment, height and MDD. Several PGS for mental illness
and personality traits are more selected for (or less against) among
women, including major depressive disorder (MDD), schizophrenia and
neuroticism, while extraversion is more selected for among men. PGS for waist
circumference and waist-hip ratio are less selected for among women, and PGS for
educational attainment are more selected against, though the difference
is only significant for the educational attainment PGS of @lee2018gene
("EA3"). One possible reason for these sex differences is
that polygenic scores may affect fertility via success in marriage markets, and
men and women may value different characteristics in these.

@@

We next focus on variables related to household type and reproductive
strategy. We split males and females by lifetime number of sexual
partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
For both sexes, selection effects are larger and more significant among
those with more than 3 partners. Next we split respondents by whether
they were living with a spouse or partner at the time of interview.
Effects are larger among those not living with a spouse or partner
(Figure \@ref(fig:plot-n-partners)b). Lastly, we split female
respondents by age at first live birth (AFLB).[^1] There is evidence for
genetic effects on AFLB [@Barban_2016], and there is a close link
between this variable and number of children born. Figure
\@ref(fig:plot-n-partners)c shows effect sizes estimated separately for
each tercile of AFLB. Several effects are strikingly different across
terciles. ADHD and MDD are selected for amongst the youngest third of
mothers, but selected against among the oldest two-thirds. Educational
attainment is selected for among the oldest two-thirds of mothers, but
is not significantly selected among the youngest third. Similarly,
several PGS of body measurements are selected against only among older
mothers. The correlation between effect sizes for the youngest and
oldest terciles is `r cor_age_flb`. As before, recall that polygenic scores may 
have two effects: changing the number of children within each category, and 
changing the number of children by shifting respondents between categories.

@@

\newgeometry{left=1cm,right=1cm,top=1cm,bottom=1cm}

@@

\restoregeometry

@@

To investigate this further, we regress number of children on PGS
*controlling* for AFLB (Figure \@ref(fig:plot-age-flb)). In
`r opp_signed` out of `r length(score_names)` cases, effects
change sign when controls are added. The correlation between effect
sizes controlling for AFLB, and raw effect sizes, is `r cor_raw_flb`.
Thus, selection effects seem to come through two opposing channels: an
effect on AFLB, and an opposite-signed effect on number of children 
controlling for AFLB.

```{r plot-age-flb, fig.cap = "Selection effects controlling for age at first live birth (women only). Effect sizes without controls (for women only) are shown for comparison."}

res_combined %>% 
      mutate(
        score_name = pretty_names(score_name),
        score_name = fct_reorder(score_name, estimate, order_abs(2)),
        Regression = ifelse(reg.type == "raw", "Uncontrolled", "Controlled for AFLB")
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(y = score_name, x = estimate, fill = Regression, 
        color = p < 0.05/{{n_regs}})) + 
        geom_vline(xintercept = 0, linetype = "dashed", color = "grey20") +
        pval_geom_point() +
        pval_scale() +
        scale_fill_brewer(type = "qual", palette = 3) +
        my_labs() +
        my_theme()
```

@@

n_regs <- as.double(length(score_names))

res_wt_combined %>% 
  mutate(
    Weights  = fct_relevel(Weights, "Geographical", "Age/Qualification",
                "Age/Qual/AFLB", "None"),
    term     = pretty_names(term),
    term     = fct_reorder(term, estimate, order_abs(1))
  ) %>% 
  rename(p = p.value) %>% 
  ggplot(aes(
    x = estimate, 
    y = term, 
    shape = Weights, 
    fill = Weights, 
    colour = p < 0.05/{{n_regs}}
  )) +
  geom_point(size = 1.5, stroke = 0.5) +
  my_vline +
  pval_scale() +
  scale_fill_manual(values = c(
    "None"              = "grey15", 
    "Geographical"      = "dodgerblue4",
    "Age/Qualification" = "steelblue2",
    "Age/Qual/AFLB"      = "darkseagreen4"
  )) +
  scale_shape_manual(values = c(
    "None" = "square filled", 
    "Geographical" = "circle filled",
    "Age/Qualification" = "circle filled", 
    "Age/Qual/AFLB" = "circle filled"
  )) + 
  labs(x = "Effect size", y = "")

@@

The UK Biobank data contains information on respondents' number of
siblings. Since respondents' polygenic scores are equal in expectation
to the mean scores of their parents, we can use this to look at selection 
effects in the parents' generation. We estimate equation (\ref{eq:regression})
using *number of siblings* (including the respondent) as the dependent variable.
The parents' generation has an additional source of ascertainment bias: sampling
parents of respondents overweights parents who have many children. For instance,
parents of three children will have, on average, three times more children
represented in UK Biobank than parents of one child. Parents of no children will
by definition not be represented. To compensate, we reweight our preferred
weightings (Age/Qualification) by the inverse of *number of siblings*.

Selection effects are highly correlated across the two generations, and most
share the same sign (Appendix Figure \@ref(fig:plot-res-sibs)). Effect size
estimates are larger for the parents' generation. We treat this result cautiously,
because when we split respondents up by year of birth, we find few differences
in effect sizes between early- and late-born respondents, for either generation.
In other words, since estimated effect sizes change when we change the dependent
variable, but do not change over time within either dependent variable, this
may be due to remaining ascertainment bias within the sample.

Although the direction of selection effects does not change between the
generations, there are other differences. Compared to residualized polygenic
scores, selection effects on unresidualized scores are about ten percent higher
on average for *n siblings*, whereas effects for *number of children* barely
change (Appendix Figure \@ref(fig:plot-regs-controlled-pcs)). This could be
because earlier fertility is driven more by geographically clustered deprivation
[e.g. via an insurance motive, @rendall1998old], which may correlate with the
broad-scale genetic variation captured by principal components, such as ancestry.

We also check whether selection effects differ by socio-economic status in the
parents' generation. We have no information about parents' income, so we use the
1971 Townsend deprivation score of respondents' birthplace as a proxy
[@townsend1987deprivation]. Results (Appendix Figure
\@ref(fig:plot-siblings-townsend)) show the same pattern as for respondents:
effect sizes are larger and more often significant in the most deprived areas.

```{r calc-siblings-children}
sib_chn_cor <- cor.test(famhist$n_sibs, famhist$n_children, use = "complete")
stopifnot(sib_chn_cor$p.value < 2e-16)
```

Lastly, the siblings data lets us check for a "quantity-quality tradeoff" between
number of children and number of grandchildren (for the parents' generation). We
do not find any: in fact, the correlation between *number of siblings* and 
*number of children* is positive ($\rho$ = `r sib_chn_cor$estimate`, 
$p < 2 \times 10^{-17}$).

@@

These results show that selection effects are weaker, absent, or even reversed
among some subgroups of the population. This can potentially be explained by the
economic theory of fertility [@becker1960economic; @willis1973new;
@becker1976child]. According to this theory, increases in a person's wage affect their
fertility via two opposing channels. There is an *income effect* by which
children become more affordable, like any other good. There is also a
*substitution effect*: since childrearing has a cost in time, the opportunity
cost of childrearing increases if one's market wage is higher. The income effect
would lead higher earners to have more children. The substitution effect would
lead higher earners to have fewer children. If so, then genetic variants which
affect earnings potential in the labour market may cause opposing effects on
fertility. The income effect will cause natural selection in favour of
earnings-increasing variants. The substitution effect will do the reverse.

@@

The theory can also explain why the natural selection pressure is weaker among
higher-income parents. @becker1976child show that if parents care about child
quality as well as quantity, under certain assumptions the income effect will be
stronger at higher income levels; this can lead to a U-shaped relationship
between income and fertility. Empirically, @cohen2013financial find that income
decreases fertility at low income levels but increases it at higher income
levels. If, in our data, at high income levels income and substitution effects
are roughly balanced , then income-linked polygenic scores will be neither
selected for nor against.

@@

Second, we re-estimate equation \@ref(eq:regression) for each
polygenic score, controlling for earnings in first job and education levels. Effect sizes
are generally reduced (Appendix Figure \@ref(fig:plot-res-earn-educ)), and only
`r n_05_33` out of 33 PGS are significant at $p = 0.05/33$. This suggests that
earnings and education indeed mediate the effect of polygenic scores on fertility.

@@

An alternative theory is that traits selected for are linked to externalizing
behaviour, risk-seeking and low time discount rates, via the channel of early
sexual behaviour [@Mills2020.05.06.081273]. The data here provide some support
for this: scores which might plausibly be linked to externalizing behaviour,
like ADHD and age at smoking initiation, are selected for. However, this theory
is less good at explaining variation in selection across the full range of
scores, including physical measures, e.g. waist-hip ratio and BMI. Externalizing
behaviour also does not explain why selection effects should work in the
opposite direction among older parents, whereas the economic theory can explain
this via income effects, as described above. We test the alternative theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040). While risk attitude is always a
highly significant predictor of number of children, it has little impact on the
effect size or significance of the PGS. The median ratio of PGS effect sizes
between regressions with and without controls is `r median(ratio)`; all PGS
which are significant at $p \le 0.05/33$ in uncontrolled regressions remain so
when controlling for risk attitude. Thus, although risk attitude does predict
number of children, it appears not to mediate selection effects. Overall, we
believe that the economic theory is the most likely explanation.

@@

Previous work has documented natural selection in modern populations on variants
underlying polygenic traits [@beauchamp2016genetic; @kong2017selection;
@sanjak2018evidence]. We show that correlations between polygenic scores and
fertility are highly concentrated among specific subgroups of the population,
including people with lower income, lower education, younger first parenthood, 
and more lifetime sexual partners. Indeed, among older parents, selection effects are
actually reversed. Furthermore, the size of selection effects on a polygenic
score correlates with that score's association with labour market earnings. The
economic theory of fertility provides a parsimonious explanation for this.

Polygenic scores which correlate with high (low) earnings and more (less) education are being
selected against (selected for). In addition, many of the phenotypes under
positive selection are linked to disease risk, or are what many people would see
as undesirable to have. For example, most people would probably prefer to have
high educational attainment, a low risk of major depressive disorder, and a low
risk of coronary artery disease, but natural selection is pushing against
genes associated with these traits. Potentially, this could increase the health
burden on modern populations, but that depends on effect sizes. Our results
suggest that naïve estimates can be affected by sample ascertainment bias. This
problem may be less serious in surveys which aim to be representative (as the UK
Biobank does not). However, there is still scope for bias, since not all
respondents consent to the collection of genetic data. For instance, completion
rates for genotype data in the US Health and Retirement Study were around 80-85%
[@hrs2020]. Researchers should be aware of the risks of ascertainment when
studying modern natural selection.

We also do not know how estimated effect sizes of natural selection will change
as more accurate measures of genetic variation are produced. And we are unsure
whether genetic variants underlying other phenotypes will show a similar pattern
of natural selection to those studied here. In addition, genetic effects on
educational attainment have been shown to be inflated in population-based
samples as compared to within-family designs, likely because of indirect genetic 
effects, gene-environment correlations, and/or assortative mating 
[@lee2018gene; @selzam2019comparing; @kong2018nature]. 
In short, it is probably too early to tell whether modern natural selection has a
substantively important effect on the genetic make-up of the population.
Nevertheless, we note that selection effects on our measured polygenic scores
are still relatively small, even after reweighting to account for ascertainment
bias.

Because selection effects are concentrated in poorer groups, they may also increase
inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
attainment (EA3), for children in households of different income groups. The
blue bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with 
educational attainment, which lowers the mean, is stronger at the bottom of the 
income distribution, and this increases the differences between groups.

@@

Any model of fertility is implicitly a model of natural selection, but so far,
the economic and human genetics literatures have developed in parallel.
Integrating the two could deepen our understanding of natural selection in 
modern societies.
Economics possesses a range of theoretical models on the effects of skills,
education and income [see @hotz1997economics; @lundberg2007american]. One
perennial problem is how to test these theories in a world where education,
labour and marriage markets all interact. Genetic data, such as polygenic
scores, could help to pin down directions of causality. Conversely, theory 
and empirical results from economics can shine a light on the mechanisms 
behind natural selection.

@@

We use participant data from UK Biobank [@bycroft2018uk], which has received
ethical approval from the National Health Service North West Centre for Research
Ethics Committee (reference: 11/NW/0382). We limit the sample to white British
participants of European descent, as defined by genetic estimated ancestry and
self-identified ethnic group [@abdellaoui2019genetic], giving a sample size of
`r comma_num(nrow(famhist))`. For regressions on number of children we use
participants over 45, since most fertility is completed by this age. This gives
a sample size of `r comma_num(nrow(famhist %>% filter(kids_ss)))`.

Polygenic scores were computed by summing the alleles across ~1.3 million 
genetic variants weighted by their effect sizes as estimated in 33 genome-wide 
association studies (GWASs) that excluded UK Biobank. To control for population 
stratification, we corrected the polygenic scores for 100 principal components (PCs). 
To compute polygenic scores and PCs, the same procedures were followed as 
described in @abdellaoui2019genetic.

@@

# Acknowledgements

Abdel Abdellaoui is supported by the Foundation Volksbond Rotterdam and by ZonMw grant 849200011 from The Netherlands Organisation for Health Research and Development. This study was conducted using UK Biobank resources under application number 40310.

\FloatBarrier
\clearpage
# Appendix

@@

\FloatBarrier
\clearpage
## Weighted regressions
